---
title: "Financial Contributions to Presidential Campaigns in Massachusetts"
author: "Debashish singh"
date: "April 17, 2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Financial Contributions to 2016 Presidential Campaigns in Massachusetts
========================================================

# Abstract 

This is an exploration of recent US presidential campaign donations in the state of Massachusetts. For this exploration data analysis, I have considered the 2016 presidential campaign finance data from [Federal Election Commission](http://fec.gov/disclosurep/PDownload.do). The dataset contains financial contribution transaction from April 18 2015 to November 24 2016. I aslo tried to do this with 2020 but we have very less an skewed data.

Throughout the analysis, I will attempt to answer the following questions:

1. Which candidate receive the most money?
2. Which candidate have the most supporters?
3. Who are those donors? What do they do?
4. How do those donors donate? Is there a pattern? If so, what is it?
5. Does Hillary Clinton receive more money from women than from men?



```{r echo=FALSE, message=FALSE, warning=FALSE}
library(gender)
library(ggplot2)
library(ggmap)
library(gridExtra)
library(dplyr)
library(lubridate)
library(zipcode)
library(aod)

```

# Univariate Plots Section


```{r echo=FALSE, warning=FALSE, message=FALSE}
ma <- read.csv('/home/debashish/Documents/Udacity Nanodegree/Term2/Project/Financial Contributions to Presidential Campaigns in Massachusetts/MA16.csv', header = TRUE, stringsAsFactors=FALSE,fill=T , sep = ",")
dim(ma)
str(ma)
```
This dataset contains 295667 contributions and 19 variables. To start with, Let's check how the contribution is distributed. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
p1 <- ggplot(aes(x = contb_receipt_amt), data = ma) +
  geom_histogram(bins = 50)
p2 <- ggplot(aes(x = 1, y = contb_receipt_amt), data = ma) +
  geom_boxplot()
grid.arrange(p1, p2, ncol = 2)
```
I observed that there were so many outliers(extreme high and extreme low values), it was impossible to see details. And there were negative contributions too. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(aes(x = contb_receipt_amt), data = ma) +
  geom_histogram(binwidth = 0.05) + 
  scale_x_log10() +
  ggtitle('Histogram of the Contribution')

tail(sort(table(ma$contb_receipt_amt)), 5)
summary(ma$contb_receipt_amt)

```
Transforming to log10 to better understand the distribution of the contribution. It looks normal and the data illustrated that most donors made small amount of contributions. 

I decided to omit the negative contributions which I believe they were refund and contributions that exceed $2700 limit, because it breaks [Federal Election Campaign Act](http://www.fec.gov/pages/fecrecord/2015/february/contriblimits20152016.shtml) and will be refunded. 
```{r}
sum(ma$contb_receipt_amt >= 2700)
sum(ma$contb_receipt_amt < 0)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Get the number of unique candidates 
unique(ma[c("cand_nm")])

# create party variables
democrat <- c("Clinton, Hillary Rodham", "Sanders, Bernard", "O'Malley, Martin Joseph", "Lessig, Lawrence", "Webb, James Henry Jr.")
ma$party <- ifelse(ma$cand_nm %in% democrat, "democrat", "republican")
ma$party[ma$cand_nm %in% c("Johnson, Gary", "McMullin, Evan", "Stein, Jill")] <- 'others'

# Get contributor's firs name for gender predication
ma$contbr_first_nm <- sub(" .*", "", sub(".*, ", "", ma$contbr_nm))

# Omit negative contributions and contributions >=2800
ma <- ma[ma$contb_receipt_amt > 0 & ma$contb_receipt_amt <= 2700, ]
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Adjust date
ma$contb_receipt_dt <- as.Date(ma$contb_receipt_dt,format = "%d-%b-%y")

# Create gender dataframe
gender_df <- gender(ma$contbr_first_nm, method = 'ssa', c(1920, 1997),
                       countries = 'United States')
gender_df

```
```{r echo=FALSE, warning=FALSE, message=FALSE}
# create gender variable
gender_df <- unique(gender_df)
names(gender_df)[1] <- 'contbr_first_nm'
ma <- inner_join(ma, gender_df, by = 'contbr_first_nm')

drops <- c('proportion_male', 'proportion_female', 'year_min', 'year_max')
ma <- ma[ , !(names(ma) %in% drops)]

```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Add missing leading zero on zipcode and clean up zipcode
ma$zip <- paste0("0", ma$contbr_zip)
ma$zip <- substr(ma$zip, 1, 5)
data(zipcode)
ma <- left_join(ma, zipcode, by = 'zip')
```
After processing the data and I have added 5 new variables to help with the analysis, and removed 5897 observations because they were either negative amount or amount exceed $2700.

The new variables are:

* party: candidates party affilliation.
* contbr_first_nm: contributor's first name will be used to predict gender.
* gender: contributor's gender.
* Latitude: Donor's latitude for map creation. 
* Longitute: Donor's longitude for map creation.

After adding the variables, I am curious what the contribution distribution looks like across the parties, candidates, genders and occupations.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Create party dataframe and bar plots for party
party_group <- group_by(ma, party)
ma.contr_by_party <- summarize(party_group,
                               sum_party = sum(contb_receipt_amt),
                               number_of_candidate = length(unique(cand_id)), 
                               mean_party = sum_party/number_of_candidate, 
                               n = n()) 
                              
ma.contr_by_party
ma.contr_by_party$party <- ordered(ma.contr_by_party$party, 
                                   levels = c('democrat', 'republican', 'others'))
ggplot(aes(x = party, y = n, fill = party), data = ma.contr_by_party) +
  geom_bar(stat = 'identity') +
  geom_text(stat = 'identity', aes(label = n),
            data = ma.contr_by_party, vjust = -0.4) +
  xlab('Party') +
  ylab('Number of Contributions') +
  ggtitle('Total Number of Contributions by Party') +
  scale_fill_manual(values = c('blue', 'red', 'gold'))
sum(ma.contr_by_party$n)
```
Till November, 2016, total number of donations made to the presidential election amounts to 269K, and the Democratic party took more than 243K and almost 10 times of the number of donations made to the Republican party. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
table(ma$cand_nm)
ggplot(aes(x = cand_nm), data = ma) + geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('candidate') +
  ylab('Number of Contributions') +
  ggtitle('Number of Contributions by Candidate')
```
There were total 25 candidates, Hillary Clinton was the leader in the number of contributions, followed by Bernard Sanders, then Donald Trump.


```{r echo=FALSE, warning=FALSE, message=FALSE}
gender_group <- group_by(ma, gender)
ma.contr_by_gen <- summarize(gender_group, 
                             sum_gen = sum(contb_receipt_amt),
                             n_gen = n())
ma.contr_by_gen
ggplot(aes(x = gender, y = n_gen, fill = gender), 
       data = ma.contr_by_gen, vjust = -0.4) +
  geom_bar(stat = 'identity') +
  geom_text(aes(label = n_gen), stat = 'identity', data = ma.contr_by_gen, vjust = -0.4) +
  xlab('Gender') +
  ylab('Number of Contributions') +
  ggtitle('Number of Contributions by Gender')
```
Surprised to know that there were a lot more women than men to made donations.

Who are those donors?

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Create occupation datadrame and occupation plots
occupation_group <- group_by(ma, contbr_occupation)
ma.contr_by_occu <- summarize(occupation_group, 
                              sum_occu = sum(contb_receipt_amt), 
                              mean_occu = mean(contb_receipt_amt), 
                              n = n())
ma.contr_by_occu <- subset(ma.contr_by_occu, contbr_occupation != "INFORMATION REQUESTED")
ma.contr_by_occu <- head(arrange(ma.contr_by_occu,desc(n)), n = 10)
ma.contr_by_occu$contbr_occupation <- ordered(ma.contr_by_occu$contbr_occupation, levels = c('RETIRED', 'NOT EMPLOYED', 'TEACHER', 'ATTORNEY', 'PROFESSOR', 'PHYSICIAN', 'CONSULTANT', 'SOFTWARE ENGINEER', 'HOMEMAKER', 'ENGINEER'))
ma.contr_by_occu
ggplot(aes(x = contbr_occupation, y = n), data = ma.contr_by_occu) +
  geom_bar(stat = 'identity') +
  xlab('Top 10 Occupations') +
  ylab('Number of Donors') +
  ggtitle('Top 10 Occupations by Number of Donors') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

When we count the number of donors, retired people take the first place, followed by unemployed, teacher and attorney. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
summary(ma$contb_receipt_dt)
ggplot(aes(x = contb_receipt_dt), data = ma) + geom_histogram(binwidth = 30, position = position_dodge()) +
  xlab('Date') +
  ylab('Number of Contributions') +
  ggtitle('Histogram of Contribution Date')
```
And it is also interesting to see when people made contributions. The date distribution appears bimodal with period peaking around March 2016 or so and again close to the election.

# Bivariate Plots Section

```{r echo=FALSE, warning=FALSE, message=FALSE}
# barplot for contribution amount
ma.contr_by_party
ggplot(aes(x = party, y = sum_party/1000, fill = party), data = ma.contr_by_party) +
  geom_bar(stat = 'identity') +
  geom_text(stat = 'identity', aes(label = round(sum_party/1000)),
            data = ma.contr_by_party, vjust = -0.4) +
  xlab('Party') +
  ylab('Contribution Received (Thousands)') +
  ggtitle('Total Contribution Amount by Party') +
  scale_fill_manual(values = c('blue', 'red', 'gold'))
ggplot(aes(x = party, y = mean_party/1000, fill = party), data = ma.contr_by_party) +
  geom_bar(stat = 'identity') +
  geom_text(stat = 'identity', aes(label = round(mean_party/1000)), 
            data = ma.contr_by_party, vjust = -0.4) +
  xlab('Party') +
  ylab('Contribution Received (Thousands)') +
  ggtitle('Average Contribution Received by Party') +
  scale_fill_manual(values = c('blue', 'red', 'gold'))
sort(by(ma$contb_receipt_amt, ma$cand_nm, sum))
ggplot(aes(x = cand_nm, y = contb_receipt_amt/1000), data = ma) + 
  geom_bar(stat = 'identity') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('candidate') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Contribution Amount by Candidate')
sum(ma$contb_receipt_amt)
```
The total contribution amount made to the presidential candidates grossed over 30 million US dollars in Massachusetts. We can easily see where the money went.

Democratic party takes the majority share of donor contribution. Democratic party got more than 25.8 mollion US dollars in total, which is approximately 6 times of what the Republican received. It is getting worse for the Republican when comes to the average amount, as there were 17 Republican candidates and only 5 Democratic candidates.

Hillary Clinton received the most contribution amount followed by Bernard Sanders then Donald Trump. 

There is no surprise as Massachusetts is the home of Kennedy family, and routinely voted for the Democratic party in federal elections. 

To see contribution patterns between parties and candidates, Let's start with boxplots.  
```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(aes(x = party, y = contb_receipt_amt, fill = party), data = ma) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 2700)) +
  xlab('party') +
  ylab('Contribution Amount') +
  ggtitle('Boxplot for Contribution Amount by Party') +
  scale_fill_manual(values = c('blue', 'gold', 'red'))
```
However, it is very hard to compare contributions among all parties due to so many outliers. I will apply log scale and remove the 'others' party from now on.

```{r echo=FALSE, warning=FALSE, message=FALSE}
ma <- subset(ma, ma$cand_nm != "McMullin, Evan" & ma$cand_nm != "Johnson, Gary" & ma$cand_nm != "Stein, Jill")
by(ma$contb_receipt_amt, ma$party, summary)
ggplot(aes(x = party, y = contb_receipt_amt, fill = party), data = ma) +
  geom_boxplot() +
  scale_y_log10() +
  xlab('party') +
  ylab('Contribution Amount') +
  ggtitle('Boxplot for Contribution Amount(Log_10) by Party') +
  scale_fill_manual(values = c('blue', 'red'))
```
Now it is much better. 

Although the Republican has the higher median and mean, the Democrat has more variations and the distribution is more spread out. This indicates that the Democrat has more big and small donors.  

```{r echo=FALSE, warning=FALSE, message=FALSE}
by(ma$contb_receipt_amt, ma$cand_nm, summary)
ggplot(aes(x = cand_nm, y = contb_receipt_amt), data = ma) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('candidate') +
  ylab('Contribution Amount') +
  ggtitle('Contribution Amount by Candidate')
```
Christopher Christie, Lindsey Graham and George Patake have the highest median, Jeb Bush has the greatest interquartile range while Hillary Clinton and Bernard Sanders seem to have the lowest median. But Hillary Clinton has the most outliers than anyone else. Bernard Sanders has significant number of outliers as well. 

Now let's check within parties.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create candidate dataframe and bar plots for candidate 
can_group <- group_by(ma, party, cand_nm)
ma.contr_by_can <- summarize(can_group, 
                             sum_can = sum(contb_receipt_amt), 
                             mean_can = mean(contb_receipt_amt),
                             n = n())
ma.contr_by_can <- arrange(ma.contr_by_can, sum_can)
ma.contr_by_can
ggplot(aes(x = cand_nm, y = sum_can/1000), data = ma.contr_by_can) +
  geom_bar(aes(fill = party), stat = 'identity') +
  scale_y_continuous(limits = c(0, 23000)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('Candidate') +
  ylab('Contribution Received (Thousands)') +
  ggtitle('Contribution Received by Candidate') +
  scale_fill_manual(values = c("blue", "red"))

# Create candidate_party dataframe
can_party <- left_join(ma.contr_by_can, ma.contr_by_party, by = 'party')
ggplot(aes(x = cand_nm, y = sum_can/sum_party*100), data = can_party) +
  geom_bar(aes(fill = party), stat = 'identity') +
  geom_text(stat='identity', aes(label = paste(round(100*sum_can/sum_party,0),'%')), 
            size=3, data = can_party, vjust = -0.4)+
  scale_y_continuous(limits = c(0, 100)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('Candidate') +
  ylab('Percentage of Donation') +
  ggtitle('Percentage of Contribution Received by Candidate from their Own Party') +
  scale_fill_manual(values = c("blue", 'red'))
```
Within each party, majority of the donations were received by only few candidates. For Democratic party, Hillary Clinton and Bernard Sanders take almost 99% of all donations to the Democratic party, and of which, 81% went to Hillary Clinton. For the Republican party, Donald Trump led the way taking 41% of all donations to the Republican party. Donald Trump, Marco Rubio, Ted Cruz, John Kasich, Jeb Bush all together taking 83% of all donations to the Republican party, the remaining 17% were shared by the other 12 Republican candidates.

From the above charts, we are able to see who were the top candidates in each party in Massachusetts. I will explore the following candidates who received at least 9% of total donations in their party in details later.

```{r echo=FALSE, warning=FALSE, message=FALSE}
top_candidate <- c("Clinton, Hillary Rodham", "Sanders, Bernard", "Trump, Donald J.", "Rubio, Marco", "Cruz, Rafael Edward 'Ted'")
top_candidate
```
We have seen earlier that women made 26% more number of contributions than men. Is that the same for the amount of money donated? And do women tend to donate more to the liberals or to woman candidate?
  
```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(aes(x = gender, y = contb_receipt_amt, fill = gender), data = ma) +
  geom_boxplot() +
  xlab('gender') +
  ylab('Contribution Amount') +
  ggtitle('Contribution Amount by Gender Boxplot') +
  coord_cartesian(ylim = c(0, 100))
by(ma$contb_receipt_amt, ma$gender, summary)
```
On average, male donated $131.12 and female donated $99.78, there is a 31% difference between genders. Female contributed much less than male when we look at median, mean and third quartile. 

```{r echo=FALSE, warning=FALSE, gender_data} 
# Create gender dataframe
gender_group <- group_by(ma, gender)
ma.contr_by_gen <- summarize(gender_group, 
                             sum_gen = sum(contb_receipt_amt),
                             n = n())
ma.contr_by_gen
ggplot(aes(x = gender, y = sum_gen/1000, fill = gender), 
       data = ma.contr_by_gen) +
  geom_bar(stat = 'identity') +
  geom_text(aes(label = sum_gen/1000), stat = 'identity', data = ma.contr_by_gen, vjust = -0.4) +
  xlab('Gender') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Contribution Amount by Gender')
```

However, when we look at the total contribution amount between genders, they were very close. 

```{r echo=FALSE, ,warning=FALSE, message=FALSE}
# Create gender_to_top_candidate dataframe for bar plot
ma.gen_to_top_candidate <- ma %>%
  filter(ma$cand_nm %in% top_candidate) %>%
  group_by(cand_nm, gender) %>%
  summarize(sum_gen_can = sum(contb_receipt_amt))
ma.gen_to_top_candidate
ggplot(aes(x = cand_nm, y = sum_gen_can/1000, fill = gender), 
       data = ma.gen_to_top_candidate) +
  geom_bar(stat = 'identity', position = position_dodge(width = 1)) +
  xlab('Candidate') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Contribution Amount to Top Candidate by Gender') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Female in Massachusetts contributed a little less than 15 million US Dollars in total to the presidential campaign in 2016, of which, more than 11 million Dollars went toward Hillary Clinton. This confirms that Massachusetts women donate more to the liberals and/or to woman candidate.

Earlier we have seen that retired people make the most number of contributions, how about total contribution amount and average contribution amount across top 10 occupations? 

```{r echo=FALSE, warning=FALSE, message=FALSE}
ma.contr_by_occu
ggplot(aes(x = contbr_occupation, y = sum_occu/1000), data = ma.contr_by_occu) +
  geom_bar(stat = 'identity') +
  geom_text(stat = 'identity', aes(label = round(sum_occu/1000)), data = ma.contr_by_occu, vjust = -0.4) +
  xlab('Top 10 Occupations') +
  ylab('Total Contribution Amount (Thousands)') +
  ggtitle('Total Contribution Amount From Top 10 Occupations') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggplot(aes(x = contbr_occupation, y = round(mean_occu,2)), data = ma.contr_by_occu) + 
  geom_bar(stat = 'identity') +
  geom_text(stat = 'identity', aes(label = round(mean_occu,2)), data = ma.contr_by_occu, vjust = -0.4) +
  xlab('Top 10 Occupations') +
  ylab('Average Contribution Amount') +
  ggtitle('Average Contributions From the Top 10 Occupations') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Retirees may make the most number of contributions, but the highest average contributions are attorneys followed by homemakes and consultant. Unemployed, being the least.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Create top_occupation datafram 
top_occu_df <- filter(ma, contbr_occupation %in% ma.contr_by_occu[['contbr_occupation']])
ggplot(aes(x = contbr_occupation, y = contb_receipt_amt), data = top_occu_df) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('Top 10 Occupations') +
  ylab('Donations Amount') +
  ggtitle('Donations Made by Top 10 Occupations')
```
I want to dive deeper to investigate the contribution amount distribution among occupations. But this one is hard to see because there are so many outliers.

```{r echo=FALSE, warning=FALSE, message=FALSE}
by(top_occu_df$contb_receipt_amt, top_occu_df$contbr_occupation, summary)
ggplot(aes(x = contbr_occupation, y = contb_receipt_amt), data = top_occu_df) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0, 200)) +
  xlab('Top 10 Occupations') +
  ylab('Donations Amount') +
  ggtitle('Donations Made by Top 10 Occupations Excl. Outliers')
```

This looks much better. After I filtered out outliers (donations that are extreme high), a boxplot confirms my above observation. The median contribution of teacher, homemaker and unemployed are relatively low.  

It is still apparent that attorney made the large contribution with the highest 
average donation and the largest variability. Some of them contributed 4 times of their respective median.

# Bivariate Analysis

### Talk about some of the interesting findings you observed in this part of the investigation.

* Majority of the total contribution in Massachusetts went towad the Democratic party.
* There were 5 Democratic candidates and 17 Republican candidates. Therefore, there is even bigger difference when we compare average amount between parties. 
* Within each party, the majority of contributions are received by a few candidates.
* In Massachusetts there are more female donors than male donors, but female donate much less than male on average.
* In Massachusetts, majority of the contributions from female donors went toward Democratic party and/or woman candidate.
* Retired people contribute the most in total amount, and software engineers and engineers are among the least in total contribution amount.
* Attorneys had the highest average contribution amount and greatest interquartile range, unemployed people have the lowest average contribution amount and one of the smallest interquartile ranges. 

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

Surprisingly, homemakers had the 2nd highest average contribution amount, but the median contribution in this group is among the lowest. It suggests that the distribution of the data is right skewed with many outliers. Also my presumption is that majority of the homemakers are women. 

### What was the strongest relationship you found?

Men had higher donation amount than women. 

# Multivariate Plots 

```{r echo=FALSE, warning=FALSE}
ma.top_candidate <- ma %>%
  filter(cand_nm %in% top_candidate) %>%
  group_by(cand_nm, contb_receipt_dt) %>%
  summarize(n = n(), total = sum(contb_receipt_amt))
ggplot(aes(x = contb_receipt_dt, y = total/1000, color = cand_nm), data = ma.top_candidate) +
  geom_jitter(alpha = 0.05) +
  geom_smooth(method = 'loess') +
  xlab('Date') +
  ylab('Contribution Amount (Thousands)') +
  ggtitle('Time Series of Contribution Amount by Candidate')
ggplot(aes(x = contb_receipt_dt, y = n, color = cand_nm), data = ma.top_candidate) +
  geom_jitter(alpha = 0.05) +
  geom_smooth(method = 'loess') +
  xlab('Date') +
  ylab('Number of Contributions') +
  ggtitle('Time Series of Number of Contributions by Candidate')
```

We know that Hillary Clinton raised the most money and had the most supporters in Massachusetts. But is this always true throughout the campaign process? When I look at above 2 graphs, I notice 2 things:

1. Bernard Sanders actually raised more money than Hillary Clinton started from January 2016 lasted for a few months.
2. Bernard Sanders actually had more supporters than Hillary Clinton from January 2016 onward until June 2016 when he announced to endorse Hillary Clinton that broke his supporters' hearts.

This only reinforces my doubt that what if Bernard Sanders would have run against Donald Trump? Even Donald Trump himself famously stated the following: [I would rather run against Crooked Hillary Clinton than Bernie. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(aes(x = contb_receipt_dt, y = total, color = cand_nm), data = ma.top_candidate) +
  geom_jitter(alpha = 0.05) +
  geom_smooth(method = 'loess') +
  xlab('Date') +
  ylab('Contribution Amount') +
  ggtitle('Time Series of Contribution Amount(Log_10) by Candidate') +
  facet_wrap(~ cand_nm) +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1))
```
Interesting to see every top candidates' time series trend. Ted Cruz had a slow and steady growth in contribution amount, that ended as soon as he suspended his campaign in May 2016. Marco Rubio dopped out even earlier in March 2016. Donald Trump's contribution donation had a steady growth until around September 2016. His campaign probably did not spend a lot of money in Massachusetts. 

As a side note, although Donald Trump did not win in Massachusetts, [A Third of Massachusetts Voters Picked Trump](http://www.bostonmagazine.com/news/blog/2016/11/10/massachusetts-trump-voters/) and [The Trump effect happened in Massachusetts, too](https://www.bostonglobe.com/metro/2016/11/13/the-trump-effect-happened-massachusetts-too/fOGkVgbSQ2LHpuixIHxi0H/story.html). 

Where do those donors reside? 

```{r echo=FALSE, warning=FALSE, message=FALSE, data_map}
# Create a new dataframe for map
lat <- ma$latitude
lon <- ma$longitude
party <- ma$party
ma_map <- data.frame(party, lat, lon)
colnames(ma_map) <- c('party', 'lat', 'lon')

# make_bbox function to get a zoom level

sbbox <- make_bbox(lon = ma$lon, lat = ma$lat, f = 0.01)
my_map <- get_map(location = sbbox, source ="google", maptype="roadmap", scale = 2, color="bw", zoom = 7)

ggmap(my_map) +
  geom_point(data=ma_map, aes(x = lon, y = lat, color = party), 
             size = 2, alpha = 0.2) +
  xlab('Longitude') +
  ylab('Latitude') +
  ggtitle('Location of Donors') +
  scale_y_continuous(limits = c(41, 43)) +
  scale_x_continuous(limits = c(-74, -70)) +
  scale_color_manual(breaks=c("democrat", "republican"), values=c("blue","red"))
```
It looks like more republicans concentrated around Boston area, this does make sense as Boston is the largest city in Massachusetts.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. 

* While closer to the election, more big pocket donors supported Hillary Clinton.
* While closer to the election, less donation went toward Donald Trump.

### Were there any interesting or surprising findings?

For a certain period of time, Bernard Sanders received more donations and gained more popularity than Hillary Clinton.

## Final Plots and Summary 

### Most Donations went toward a few candidates.

```{r echo=FALSE, warning=FALSE, message=FALSE}
can_party$cand_nm <- ordered(can_party$cand_nm, levels = c("Clinton, Hillary Rodham", "Sanders, Bernard" , "O'Malley, Martin Joseph", "Lessig, Lawrence", "Trump, Donald J.", "Rubio, Marco", "Cruz, Rafael Edward 'Ted'", "Kasich, John R.", "Bush, Jeb", "Carson, Benjamin S.", "Christie, Christopher J.", "Graham, Lindsey O.", "Fiorina, Carly", "Paul, Rand", "Walker, Scott", "Huckabee, Mike", "Santorum, Richard J.", "Pataki, George E.", "Gilmore, James S III", "Perry, James R. (Rick)", "Jindal, Bobby", "Webb, James Henry Jr."))
ggplot(aes(x = cand_nm, y = sum_can/sum_party*100), data = can_party) +
  geom_bar(aes(fill = party), stat = 'identity') +
  geom_text(stat='identity', aes(label = paste(round(100*sum_can/sum_party,0),'%')), 
            size=3, data = can_party, vjust = -0.4)+
  scale_y_continuous(limits = c(0, 100)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('Candidate') +
  ylab('Percentage of Donation') +
  ggtitle('Percentage of Contribution Received by Candidate from Their Own Party') +
  scale_fill_manual(values = c("blue", 'red'))
```

In Massachusetts, the financial donations to the presidential campaign were distributed unevenly. Especially in Democrat, 99% of the donations for Democrat went to two candidates and Hillary Clinton took 81%. It is obvious that Massachusetts is among the bluest of states. 

## Contribution by Occupation

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(aes(x = contbr_occupation, y = sum_occu/1000), data = ma.contr_by_occu) +
  geom_bar(stat = 'identity') +
  geom_text(stat = 'identity', aes(label = round(sum_occu/1000)), data = ma.contr_by_occu, vjust = -0.4) +
  xlab('Top 10 Occupations') +
  ylab('Total Contribution Amount (Thousands)') +
  ggtitle('Total Contribution Amount From Top 10 Occupations') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The total contribution across occupations differ substantially. If I was asked which occupation contributed the most to presidential candidates in Massachusetts in 2016, I would have guessed 'attorneys' or 'CEOs'. Wrong.

The top occupation isn't really an 'occupation', but individuals who lists their occupation as 'retired' in federal documents. Unlike attorneys or consultants whose donations may covered by their companies, retired people more likely pay from their own pockets. 

It continues to surprise me that software engineer among the lowest in total contribution, considering their presumably above average salary. But any further conclusion requires better knowledge of industry political background.

### Time Series of Top Candidates
```{r echo=FALSE, warning=FALSE, message=FALSE}
p1 <- ggplot(aes(x = contb_receipt_dt, y = total, color = cand_nm), 
             data = ma.top_candidate) +
  geom_jitter(alpha = 0.05) +
  geom_smooth(method = 'loess') +
  scale_y_log10() +
  xlab('Date') +
  ylab('Contribution Amount') +
  ggtitle('Time Series of Contribution Amount(Log_10) by Candidate')
p2 <- ggplot(aes(x = contb_receipt_dt, y = n, color = cand_nm), 
             data = ma.top_candidate) +
  geom_jitter(alpha = 0.05) +
  geom_smooth(method = 'loess') +
  scale_y_log10() +
  xlab('Date') +
  ylab('Number of Contributions') +
  ggtitle('Time Series of Number of Contributions(Log_10) by Candidate')
grid.arrange(p1, p2, ncol = 1)
```


Hillary Clinton dominated the contribution amount and number of contributions, the closer to the election, the more supporters with more money came to her. 

On the other hand, Bernard Sanders had a steady growth in terms of donation amount and number of donors, until he gave up his run. 

## Reflection

### Challenges and Struggles

The original Massachusetts 2016 presidential campaign contributions data contains over 295000 entries from April 2015 until November 2016. Throughout the analysis, I dealt with several issues:

* The original dataset did not contain gender information, to analyze the relationship between gender and donations, I added gender column using R's gender package which used to predict gender from donor's first name.

* To see a better picture of donors' geographic location, I added latitude and longitude columns using zipcode package and I was able to create a map using ggmap after that. 

* I chose to omit negative contributions and contributions that exceed $2700 because of the Contribution Limits for 2015-2016 Federal Elections. 

* I am not familar with ggmap and spent a lot of time on them.

### Success

* The ggplot2 and dplyr packages are the most important packages for this project. I also learned gender and zipcode packages and found they are powerful. 

* I learned a lot of new things throughout this project. Thanks to [ggmap quickstart](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf), [R-bloggers](https://www.r-bloggers.com/how-to-perform-a-logistic-regression-in-r/) and [Logistic Regression in R tutorial](https://www.youtube.com/watch?v=mteljf020EE) to make my project possible. It was a great experience. 

### Conclusion

By analyzing Massachusetts financial donation data, I found several interesting characteristics: 

* It is no doubt that Massachusetts is one of the bluest states.
* Few candidates collected the most donations. 
* Female tend to donate more to liberals and/or to female candidate.
* The retired people are the largest contribution group, and software engineers make very small contributions.
* Bernard Sanders gained more popularity than Hillary Clinton until he gave up his run.

### Future Work

The analysis I conducted is for Massachusetts state only. It would be interesting to analyze campaign finance data for more states, as well as campaign finance data nationwide. I am sure we will see things differently as data tends to surprise us. 

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
